### 2024-03-27 thesis/sections/6.1-object-identification.tex
learning curves, architectures

\subsubsection*{Results}
When looking at the results of the experiments, it can be seen that the agents are successfully discriminating the objects across all datasets.
Tables \ref{tab:results_discriminator_dale-2} to \ref{tab:results_discriminator_color} give an overview over the accuracy scores of selected configurations.
For each dataset there are many configurations that outperform the baseline models by a high margin.
However, there are still clear differences between the accuracy scores of each dataset.

\cmtDK[inline]{baseline}
\cmtDK[inline]{learning curve}
% SD: As we discussed at Semdial, it would be good to have a learning curves here that would show how fast or slow the languages converge. As David pointed out it may be that convergence is unstable and changes happen in between.
% DK: TODO

>>>>

\subsubsection*{Results}

\begin{figure}[ht!]
    \centering
    \subfigure[\emph{Dale-2} dataset with different $|V|$ highlighted]{
        \includegraphics[width=0.46\linewidth]{figures/learning-curve_discriminator_dale-2_vocab-size.png}
        \label{fig:learning-curve_discriminator_dale-2_vocab-size}
    }
    \subfigure[\emph{Dale-2} dataset with different $n$ highlighted]{
        \includegraphics[width=0.46\linewidth]{figures/learning-curve_discriminator_dale-2_max-len.png}
        \label{fig:learning-curve_discriminator_dale-2_max-len}
    }
    \subfigure[\emph{Dale-5} dataset with different $|V|$ highlighted]{
        \includegraphics[width=0.46\linewidth]{figures/learning-curve_discriminator_dale-5_vocab-size.png}
        \label{fig:learning-curve_discriminator_dale-5_vocab-size}
    }
    \subfigure[\emph{Dale-5} dataset with different $n$ highlighted]{
        \includegraphics[width=0.46\linewidth]{figures/learning-curve_discriminator_dale-5_max-len.png}
        \label{fig:learning-curve_discriminator_dale-5_max-len}
    }
    \subfigure[\emph{CLEVR color} dataset with different $|V|$ highlighted]{
        \includegraphics[width=0.46\linewidth]{figures/learning-curve_discriminator_colour_vocab-size.png}
        \label{fig:learning-curve_discriminator_colour_vocab-size}
    }
    \subfigure[\emph{CLEVR color} dataset with different $n$ highlighted]{
        \includegraphics[width=0.46\linewidth]{figures/learning-curve_discriminator_colour_max-len.png}
        \label{fig:learning-curve_discriminator_colour_max-len}
    }
    \caption{Learning curves of all discrimination games on each dataset. In the left column, the colors correspond to different vocabulary sizes $|V|$ while in the right column, the colors correspond to different message lengths $n$. The baseline is marked in black.}
    \label{fig:learning-curves_discriminator}
\end{figure}

When looking at the results of the experiments, it can be seen that the agents are successfully discriminating the objects across all datasets.
Tables \ref{tab:results_discriminator_dale-2} to \ref{tab:results_discriminator_color} give an overview over the accuracy scores of selected configurations.
For each dataset there are many configurations that outperform the baseline models by a high margin.
However, there are still clear differences between the accuracy scores of each dataset.

\cmtDK[inline]{learning curve}
% SD: As we discussed at Semdial, it would be good to have a learning curves here that would show how fast or slow the languages converge. As David pointed out it may be that convergence is unstable and changes happen in between.
% DK: TODO

### 2024-03-27 thesis/sections/6.2-re-generation.tex
learning curves, architectures

In a next step, it is tested if the agents can learn to extract features of the objects together.
For this the same sender as in the previous section is used.
Based on the results, $h_s$ is fixed to $500$ and $e_s$ to $100$
However, the task and by that the receiver's architecture is adapted in two different ways.
In the first setup, the \textbf{referring expression generator}, the receiver is tasked to describe the target object with natural language following the \emph{incremental GRE-algorithm} based on the sender's message (see Figure \ref{fig:caption_generator_game_architecture}).
This follows the approach of the approach of the \emph{RE generators} described in section \ref{sec:referring_expression_generation}.
The whole scene is encoded using the \emph{image encoder} submodule, projecting it to the image encoding dimension $e_{ri}=100$.
The sender's message is decoded with the hidden size $h_r=500$ and $e_r=100$, and concatenated with the encoded image.
This is then reduced to $LSTM_o=1500$ dimensions based on the results of the previous experiments, and used as the initial state of the captioning LSTM.
Tokens are embedded with $LSTM_e=15$ dimensions.
Since the position of the padding as well as the order of the words in the target didn't have an effect on the results, the usual approach in natural generation task is used: padding tokens are appended and the referring expression is not reversed.
During training, the ground truth caption is used as the input to the LSTM using teacher forcing.
% SD: I'm not sure how teacher forcing is implemented, see my earlier comment
% DK: typo, teacher forcing is actually applied (done)
When presented with test data, the LSTM always produces three tokens, by using its own predicted words as the input for the next step.
The loss is calculated using cross entropy.

\begin{figure}[ht]
    \centering
    \includegraphics[width=.7\linewidth]{figures/arch_caption_generator_game.png}
    \caption{Simplified architecture of the caption generator game}
    \label{fig:caption_generator_game_architecture}
    % SD: But here we don't need an LSTM, we just want to identify one of the objects. The LSTM should only encode the inout message.
    % DK: TODO
\end{figure}

However, this setup could lead to two problems:
First, the task is quite complex for both agents to learn, since two language generation steps are involved.
This might stop the agents from converging towards a useful language.
Secondly, in case a language emerges, it could be aligned with the natural language of the target referring expressions.
In other words, the sender might just learn to effectively repeat the target referring expressions without extracting the necessary features themselves.
For that reason, in the second setup the receiver is tasked to predict only the attributes of the target object in the shape of a one-hot vector.
By this, no natural language including its salience order is involved and the complexity of the complete game is reduced.
The receiver uses the same way to encode and combine the complete scene and the sender's message with $e_{ri}=100$, $h_r=500$ and $e_r=100$.
This is then passed to a linear layer that produces a vector with 13 dimension, corresponding to the 2 sizes, 3 shapes, and 8 colors.
The loss is then calculated using binary cross entropy.

\cmtDK[inline]{figure}

@@

Since the agents are trained to describe the target object discriminatively based on the described GRE-algorithm, they are trained on the \emph{Dale-2}, \emph{Dale-5} and \emph{CLEVR color} dataset.
The \emph{Dale-5} and the \emph{CLEVR color} should be again much harder to learn, since there are more objects that the agents need to discriminate the target object from.
% SD: So the sender is generating descriptions following the GRE policy?
% DK: TODO
The same metrics as in the section \ref{sec:referring_expression_generation} are used to evaluate the results for the first setup.
When predicting one-hot vectors, an overall \textbf{accuracy} is reported that reflects if all attributes were predicted correctly.
Additionally, the accuracy for each attribute is calculated separately to show which attributes give bigger challenges for the models.

\subsubsection*{Results}

@@

@@

Finally, Table \ref{tab:results:attribute-predictor-game} shows the results for the games in which the receiver predicts a one-hot vector of the attributes.
Again, the baselines are games, the receiver plays alone.
Hereby, it can be seen that the receiver can already predict the \emph{shape} and \emph{color}  very well: 88,91\% and 92\% respectively on the \emph{Dale-2} dataset, 80,78\% and 89,38\% respectively on the \emph{Dale-5} dataset, and 100\% and 90,78\% respectively on the \emph{CLEVR color} dataset.
The \emph{size} however corresponds to a random guess on the \emph{Dale} dataset and only achieves 76,72\% on the \emph{CLEVR color} dataset.
This shows that the receiver can already identify attributes of the target object on their own and discriminate it from the distractors.
However, the size still seems to be an attribute which is difficult to extract from the visual scene.
The increase in performance compared to the generation of referring expression can be explained by two reasons:
First, the complexity of the receiver is greatly reduced.
When only predicting one-hot vectors, much fewer layers and weights are involved that need to be trained and hence the models can learn the given task.
Secondly, the complexity of the task is reduced as well.
The generation of referring expressions not only includes the learning of the tokens (and their embeddings), but also their order in the referring expression and their salience for the incremental algorithm.
Using one-hot vectors, both order and salience are irrelevant and only the attributes themselves are needed to be learned.
The goal of these experiments is now to see if the information of the sender improves the predictions of the receiver.

@@

Using the \emph{CLEVR color} dataset, almost no improvements can be seen.
The predictions of the \emph{shape} stay at 100\% accuracy, while the predictions of the \emph{color} get slightly worse for most configurations and if they improve they only improve by maximum 1\% point.
The predictions of the \emph{size} stay the same.
The fact that the sender has a smaller influence on the \emph{CLEVR color} dataset than on the \emph{Dale} dataset is surprising since the sender would need to communicate only the \emph{color} instead of additionally the \emph{shape} (and \emph{size}).
A reason why this happens might be the larger number of distractors.
This number might hinder the receiver to converge and understand the task, by connecting the one-hot vector to the colors of the target object.
While for the \emph{Dale} datasets the receiver has a choice between only 2 to 5 objects, they need to decide between up to 10 objects on the \emph{CLEVR color} dataset.
That might confuse the model where to focus on and subsequently to connect it to the on-hot vector.
Future experiments with a similar dataset including fewer distractors would need to verify this hypothesis.

>>>>

In a next step, it is tested if the agents can learn to extract features of the objects together.
For this the same sender as in the previous section is used.
Based on the results, $h_s$ is fixed to $500$ and $e_s$ to $100$.
However, the task and by that the receiver's architecture is adapted in two different ways.
In the first setup, the \textbf{referring expression generator}, the receiver is tasked to describe the target object with natural language following the \emph{incremental GRE-algorithm} based on the sender's message (see Figure \ref{fig:caption_generator_game_architecture}).
This follows the approach of the approach of the \emph{RE generators} described in section \ref{sec:referring_expression_generation}.
The whole scene is encoded using the \emph{image encoder} submodule, projecting it to the image encoding dimension $e_{ri}=100$.
The sender's message is decoded with the hidden size $h_r=500$ and $e_r=100$, and concatenated with the encoded image.
This is then reduced to $LSTM_o=1500$ dimensions based on the results of the previous experiments, and used as the initial state of the captioning LSTM.
Tokens are embedded with $LSTM_e=15$ dimensions.
Since the position of the padding as well as the order of the words in the target didn't have an effect on the results, the usual approach in natural generation task is used: padding tokens are appended and the referring expression is not reversed.
During training, the ground truth caption is used as the input to the LSTM using teacher forcing.
% SD: I'm not sure how teacher forcing is implemented, see my earlier comment
% DK: typo, teacher forcing is actually applied (done)
When presented with test data, the LSTM always produces three tokens, by using its own predicted words as the input for the next step.
The loss is calculated using cross entropy.

\begin{figure}[ht]
    \centering
    \includegraphics[width=.7\linewidth]{figures/arch_caption_generator_game.png}
    \caption{Simplified architecture of the referring expression generator game}
    \label{fig:caption_generator_game_architecture}
    % SD: But here we don't need an LSTM, we just want to identify one of the objects. The LSTM should only encode the inout message.
    % DK: TODO
\end{figure}

However, this setup could lead to two problems:
First, the task is quite complex for both agents to learn, since two language generation steps are involved.
This might stop the agents from converging towards a useful language.
Secondly, in case a language emerges, it could be aligned with the natural language of the target referring expressions.
In other words, the sender might just learn to effectively repeat the target referring expressions without extracting the necessary features themselves.
For that reason, in the second setup, the \textbf{attribute generator} the receiver is tasked to predict only the attributes of the target object in the shape of a multi-hot vector (shown in Figure \ref{fig:attribute_generator_game_architecture}).
By this, no natural language including its salience order is involved and the complexity of the complete game is reduced.
The receiver uses the same way to encode and combine the complete scene and the sender's message with $e_{ri}=100$, $h_r=500$ and $e_r=100$.
This is then passed to a linear layer that produces a vector with 13 dimension, corresponding to the 2 sizes, 3 shapes, and 8 colors.
The loss is then calculated using binary cross entropy.

\begin{figure}[ht]
    \centering
    \includegraphics[width=.55\linewidth]{figures/arch_attribute_generator_game.png}
    \caption{Simplified architecture of the attribute generator game}
    \label{fig:attribute_generator_game_architecture}
\end{figure}

@@

Since the agents are trained to describe the target object discriminatively based on the described GRE-algorithm, they are trained on the \emph{Dale-2}, \emph{Dale-5} and \emph{CLEVR color} dataset.
The \emph{Dale-5} and the \emph{CLEVR color} should be again much harder to learn, since there are more objects that the agents need to discriminate the target object from.
% SD: So the sender is generating descriptions following the GRE policy?
% DK: TODO
The same metrics as in the section \ref{sec:referring_expression_generation} are used to evaluate the results for the first setup.
When predicting multi-hot vectors, an overall \textbf{accuracy} is reported that reflects if all attributes were predicted correctly.
Additionally, the accuracy for each attribute is calculated separately to show which attributes give bigger challenges for the models.

\subsubsection*{Results}
\begin{figure}[ht!]
    \centering
    \subfigure[\emph{Dale-2} dataset with different $|V|$ highlighted]{
        \includegraphics[width=0.46\linewidth]{figures/learning-curve_re-generator_dale-2_vocab-size.png}
        \label{fig:learning-curve_re-generator_dale-2_vocab-size}
    }
    \subfigure[\emph{Dale-2} dataset with different $n$ highlighted]{
        \includegraphics[width=0.46\linewidth]{figures/learning-curve_re-generator_dale-2_max-len.png}
        \label{fig:learning-curve_re-generator_dale-2_max-len}
    }
    \subfigure[\emph{Dale-5} dataset with different $|V|$ highlighted]{
        \includegraphics[width=0.46\linewidth]{figures/learning-curve_re-generator_dale-5_vocab-size.png}
        \label{fig:learning-curve_re-generator_dale-5_vocab-size}
    }
    \subfigure[\emph{Dale-5} dataset with different $n$ highlighted]{
        \includegraphics[width=0.46\linewidth]{figures/learning-curve_re-generator_dale-5_max-len.png}
        \label{fig:learning-curve_re-generator_dale-5_max-len}
    }
    \caption{Learning curves of all referring expression generation games on the \emph{Dale} datasets. In the left column, the colors correspond to different vocabulary sizes $|V|$ while in the right column, the colors correspond to different message lengths $n$. The baseline is marked in black.}
    \label{fig:learning-curves_re-generator}
\end{figure}

@@

\begin{figure}[ht!]
    \centering
    \subfigure[\emph{Dale-2} dataset with different $|V|$ highlighted]{
        \includegraphics[width=0.46\linewidth]{figures/learning-curve_oh-generator_dale-2_vocab-size.png}
        \label{fig:learning-curve_oh-generator_dale-2_vocab-size}
    }
    \subfigure[\emph{Dale-2} dataset with different $n$ highlighted]{
        \includegraphics[width=0.46\linewidth]{figures/learning-curve_oh-generator_dale-2_max-len.png}
        \label{fig:learning-curve_oh-generator_dale-2_max-len}
    }
    \subfigure[\emph{Dale-5} dataset with different $|V|$ highlighted]{
        \includegraphics[width=0.46\linewidth]{figures/learning-curve_oh-generator_dale-5_vocab-size.png}
        \label{fig:learning-curve_oh-generator_dale-5_vocab-size}
    }
    \subfigure[\emph{Dale-5} dataset with different $n$ highlighted]{
        \includegraphics[width=0.46\linewidth]{figures/learning-curve_oh-generator_dale-5_max-len.png}
        \label{fig:learning-curve_oh-generator_dale-5_max-len}
    }
    \caption{Learning curves of all attribute-predictor games on the \emph{Dale} datasets. The graphs show the accuracy of the \emph{shape}, since this highlights the learning effect best. In the left column, the colors correspond to different vocabulary sizes $|V|$ while in the right column, the colors correspond to different message lengths $n$. The baseline is marked in black.}
    \label{fig:learning-curves_oh-generator}
\end{figure}

@@

Finally, Table \ref{tab:results:attribute-predictor-game} shows the results for the games in which the receiver predicts a multi-hot vector of the attributes.
Again, the baselines are games, the receiver plays alone.
Hereby, it can be seen that the receiver can already predict the \emph{shape} and \emph{color}  very well: 88,91\% and 92\% respectively on the \emph{Dale-2} dataset, 80,78\% and 89,38\% respectively on the \emph{Dale-5} dataset, and 100\% and 90,78\% respectively on the \emph{CLEVR color} dataset.
The \emph{size} however corresponds to a random guess on the \emph{Dale} dataset and only achieves 76,72\% on the \emph{CLEVR color} dataset.
This shows that the receiver can already identify attributes of the target object on their own and discriminate it from the distractors.
However, the size still seems to be an attribute which is difficult to extract from the visual scene.
The increase in performance compared to the generation of referring expression can be explained by two reasons:
First, the complexity of the receiver is greatly reduced.
When only predicting multi-hot vectors, much fewer layers and weights are involved that need to be trained and hence the models can learn the given task.
Secondly, the complexity of the task is reduced as well.
The generation of referring expressions not only includes the learning of the tokens (and their embeddings), but also their order in the referring expression and their salience for the incremental algorithm.
Using multi-hot vectors, both order and salience are irrelevant and only the attributes themselves are needed to be learned.
The goal of these experiments is now to see if the information of the sender improves the predictions of the receiver.

@@

Using the \emph{CLEVR color} dataset, almost no improvements can be seen.
The predictions of the \emph{shape} stay at 100\% accuracy, while the predictions of the \emph{color} get slightly worse for most configurations and if they improve they only improve by maximum 1\% point.
The predictions of the \emph{size} stay the same.
The fact that the sender has a smaller influence on the \emph{CLEVR color} dataset than on the \emph{Dale} dataset is surprising since the sender would need to communicate only the \emph{color} instead of additionally the \emph{shape} (and \emph{size}).
A reason why this happens might be the larger number of distractors.
This number might hinder the receiver to converge and understand the task, by connecting the multi-hot vector to the colors of the target object.
While for the \emph{Dale} datasets the receiver has a choice between only 2 to 5 objects, they need to decide between up to 10 objects on the \emph{CLEVR color} dataset.
That might confuse the model where to focus on and subsequently to connect it to the on-hot vector.
Future experiments with a similar dataset including fewer distractors would need to verify this hypothesis.

### 2024-03-27 thesis/sections/6.3-reference-resolution.tex
learning curves, architectures

\begin{figure}[ht]
    \centering
    \includegraphics[width=.7\linewidth]{figures/arch_coordinate_predictor_game.png}
    \caption{Simplified architecture of the masked coordinate predictor game}
    \label{fig:coordinate_predictor_game_architecture}
    % SD: Again, why LSTM, are we inceremtally identifying the image while we are processing each input word?
    % DK: because it's a message sequence of discrete symbols (done)
\end{figure}

Secondly, in the \textbf{attention reference resolver}, the agents learn to point towards regions in the image, which is based on the \emph{attention reference resolver} described in section \ref{sec:reference_resolution}.
As before, the LSTM is replaced with the \emph{EGG} LSTM to parse the sender's message with the hidden size $h_r=500$ and $e_r=100$.
The dot product is calculated between each encoded region of the image and the encoded message, and the $softmax$ function is applied subsequently.
The loss is calculated using \emph{binary cross entropy} between the predictions and ground truth regions.

\cmtDK[inline]{figure}

@@

\subsection*{Results}

@@

>>>>

\begin{figure}[ht]
    \centering
    \includegraphics[width=.7\linewidth]{figures/arch_coordinate_predictor_game.png}
    \caption{Simplified architecture of the coordinate predictor game}
    \label{fig:coordinate_predictor_game_architecture}
    % SD: Again, why LSTM, are we inceremtally identifying the image while we are processing each input word?
    % DK: because it's a message sequence of discrete symbols (done)
\end{figure}

Secondly, in the \textbf{attention reference resolver}, the agents learn to point towards regions in the image, which is based on the \emph{attention reference resolver} described in section \ref{sec:reference_resolution}.
As before, the LSTM is replaced with the \emph{EGG} LSTM to parse the sender's message with the hidden size $h_r=500$ and $e_r=100$.
The dot product is calculated between each encoded region of the image and the encoded message, and the $softmax$ function is applied subsequently.
Figure \ref{fig:attention_predictor_game_architecture} shows the architecture.
The loss is calculated using \emph{binary cross entropy} between the predictions and ground truth regions.

\begin{figure}[ht]
    \centering
    \includegraphics[width=.7\linewidth]{figures/arch_attention_predictor_game.png}
    \caption{Simplified architecture of the attention predictor game}
    \label{fig:attention_predictor_game_architecture}
    % SD: Again, why LSTM, are we inceremtally identifying the image while we are processing each input word?
    % DK: because it's a message sequence of discrete symbols (done)
\end{figure}

@@

\subsection*{Results}
\begin{figure}[ht!]
    \centering
    \subfigure[\emph{Dale-2} dataset]{
        \includegraphics[width=0.46\linewidth]{figures/learning-curve_coordinate-predictor_dale-2.png}
        \label{fig:learning-curve_coordinate-predictor_dale-2}
    }
    \subfigure[\emph{Dale-5} dataset]{
        \includegraphics[width=0.46\linewidth]{figures/learning-curve_coordinate-predictor_dale-5.png}
        \label{fig:learning-curve_coordinate-predictor_dale-5}
    }
    \subfigure[\emph{CELVR color} dataset]{
        \includegraphics[width=0.46\linewidth]{figures/learning-curve_coordinate-predictor_colour.png}
        \label{fig:learning-curve_coordinate-predictor_colour}
    }
    \caption{Learning curves of all discrimination games on each dataset. As almost no games beat the baseline, $|V|$ and $n$ are not highlighted separately. The baseline is marked in black.}
    \label{fig:learning-curves_coordinate-predictor}
\end{figure}

@@

\begin{figure}[ht!]
    \centering
    \subfigure[\emph{Dale-2} dataset with different $|V|$ highlighted]{
        \includegraphics[width=0.46\linewidth]{figures/learning-curve_attention-predictor_dale-2_vocab-size.png}
        \label{fig:learning-curve_attention-predictor_dale-2_vocab-size}
    }
    \subfigure[\emph{Dale-2} dataset with different $n$ highlighted]{
        \includegraphics[width=0.46\linewidth]{figures/learning-curve_attention-predictor_dale-2_max-len.png}
        \label{fig:learning-curve_attention-predictor_dale-2_max-len}
    }
    \subfigure[\emph{Dale-5} dataset with different $|V|$ highlighted]{
        \includegraphics[width=0.46\linewidth]{figures/learning-curve_attention-predictor_dale-5_vocab-size.png}
        \label{fig:learning-curve_attention-predictor_dale-5_vocab-size}
    }
    \subfigure[\emph{Dale-5} dataset with different $n$ highlighted]{
        \includegraphics[width=0.46\linewidth]{figures/learning-curve_attention-predictor_dale-5_max-len.png}
        \label{fig:learning-curve_attention-predictor_dale-5_max-len}
    }
    \subfigure[\emph{CLEVR color} dataset with different $|V|$ highlighted]{
        \includegraphics[width=0.46\linewidth]{figures/learning-curve_attention-predictor_colour_vocab-size.png}
        \label{fig:learning-curve_attention-predictor_colour_vocab-size}
    }
    \subfigure[\emph{CLEVR color} dataset with different $n$ highlighted]{
        \includegraphics[width=0.46\linewidth]{figures/learning-curve_attention-predictor_colour_max-len.png}
        \label{fig:learning-curve_attention-predictor_colour_max-len}
    }
    \caption{Learning curves of all discrimination games on each dataset. In the left column, the colors correspond to different vocabulary sizes $|V|$ while in the right column, the colors correspond to different message lengths $n$. The baseline is marked in black.}
    \label{fig:learning-curves_attention-predictor}
\end{figure}
